Heuristics and Really Hard Instances for Subgraph Isomorphism Problems

### 2016-01-13 paper/paper.tex
State space
More outline

@@

@@

Some stuff about the complexity peak in the Glasgow solver. This looks remarkably similar to random
3SAT problems---compare, for example, Figure 1 of \citet{LeytonBrown:2014}. In particular,
satisfiable instances tend to be easier but show greater variation than unsatisfiable instances, and
there are exceptionally hard satisfiable instances \citep{Smith:1997}.

@@

What if we alter the edge probabilities for both the pattern graph and the target graph?  In the top
row of \cref{figure:non-induced} we show the phase transition for the non-induced variant, for
patterns of order 10, 20 and 30, a target of order 150, and varying pattern (x-axis) and target
(y-axis) edge probabilities. Inside the orange region, at the bottom right of each plot, every
instance is unsatisfiable---here we are trying to find a dense pattern in a sparse target. In the
purple region, at the top left, every instance is satisfiable---we are looking for a sparse pattern
in a dense target (which is easy, since we only have to preserve adjacency, not non-adjacency). The
white band between the regions shows the location of the phase transition: here, roughly half the
instances are satisfiable. (We discuss the black line below.)

On subsequent rows, we show the average difficulty of different algorithms.  We measure search
nodes. Cannot be used to compare performance of algorithms directly, but this does show that each
solver found the same set of instances difficult.  Satisfiable is easy, until very close to the
phase transition. Unsatisfiable is much harder than satisfiable. This is solver-independent,
although VF2 is terrible.

@@

To predict the location of the phase transition, we can calculate the expected number of solutions
by making a few dubious assumptions involving rounding and independence, and drawing inspiration
from the association graph encoding. Since we are trying to find an \emph{injective} mapping from a
pattern $P = G(p, d_p)$ to a target $T = G(t, d_t)$, there are $t^{\underline{p}} = t \cdot (t -
1) \cdot \ldots \cdot (t - p + 1)$ possible assignments of target vertices to pattern vertices.  We expect the
pattern to have $d_p \cdot \binom{p}{2}$ edges, so we obtain the probability of all of these edges
being mapped to edges in the target by raising $d_t$ to this power, giving an expected number of
solutions of \[ \langle Sol \rangle = t^{\underline{p}} \cdot {d_t}^{d_p \cdot \binom{p}{2}}
\textnormal{.} \] This formula predicts a very sharp phase transition from $\langle Sol \rangle \ll
1$ to $\langle Sol \rangle \gg 1$. We plot where this occurs using black lines in the first row of
\cref{figure:non-induced}.

@@

@@

?? Note that the Glasgow heuristics are moving us towards the satisfiable region, or increasing the
expected number of solutions \citep{Gent:1996:EN}.

?? LAD does SDF with no tiebreaking, no value ordering heuristic.

@@

?? I suppose we should mention kappa \citep{Gent:1996:Kappa}, etc, here. This whole thing is a bit
subtle: maximising E(N) or Kappa as a principle is encoding-independent, but whether this is done by
SDF, deg, etc, is encoding-dependent.

@@

In \cref{figure:induced} we show the induced behaviour (modified Glasgow to include the complement
as a supplemental graph).

Pattern order 10, what we expect. Patterns 20 and 30 have large unsatisfiable regions in the middle,
but these remain hard. Pattern orders 14, 15 and 16 show the transition between the two styles.

@@

?? This suggests that we're not able to make use of adjacency and non-adjacency simultaneously. Note
that this is solver-independent. In the following section we see whether this is an oddity of
subgraph isomorphism algorithms.

Odd unsat with LAD at the top right (not EHPs), trivial for Glasgow: mapping a non-clique into a
clique. Glasgow looks at non-edges and can see immediately that no mapping is possible.

@@

?? Explain why we might guess degree heuristics just won't work for induced.

?? Explain bottom row of \cref{figure:induced}, and give way of calculating it.

@@

@@

\begin{figure*}[t]
    \input{gen-graph-sat.tex}
    \caption{Behaviour of other solvers on the induced variant on smaller graphs, shown in the style of
        \cref{figure:non-induced}. The second row shows the number of search nodes used by the
    Glasgow algorithm, the third row shows the number of decisions made by the Clasp pseudo-boolean
solver, and the fourth row shows the number of search nodes used by a clique
encoding.}\label{figure:alt}
\end{figure*}

?? The odd stuff in the middle, which isn't near a phase transition, but is hard.

Is this just a weakness of current subgraph isomorphism solvers? In \cref{figure:alt} we repeat the
experiments on smaller pattern and target graphs, using different solving techniques. Although these
techniques are not competitive in absolute terms, we wish to see if the same pattern of behaviour
occurs.

Direct pseudo-boolean encoding: variables for each assignment, cardinality constraints, edge and
non-edge constraints. We use the Clasp solver. We see that hard instances remain hard, including
inside the central region, and that the easy satisfiable instances remain easy.

@@

Clique encoding: association graph, a version of BBMC \citep{SanSegundo:2011} modified to solve the
decision problem. The hard instances in the central region remain hard. Additionally, some of the
easy satisfiable instances become hard. Possible weakness in variable ordering heuristics.

Some easy instances become hard, here, but hard instances remain hard. In particular, the large
central region for the larger pattern graphs which is not near the phase transitions is still
relatively difficult in all encodings. This suggests that these instances may genuinely be hard,
despite not being near a phase transition, and that this does not indicate a weakness in subgraph
isomorphism solvers.

@@

Dynamic heuristics

Other random models and non-random

Does not give hard instances for graph isomorphism

@@

>>>>

?? Actually, it's not necessarily obvious that these are the right parameters to use... For example,
mean degree is a meaningful parameter for random graphs... Also, the formula shows it's not exactly
the ``relative'' order, but some other function of the two orders.

@@

?? I'm sure I figured out how to make overlines typeset better. Remember to look this up.

@@

?? Some stuff about the complexity peak in the Glasgow solver. This looks remarkably similar to random
3SAT problems---compare, for example, Figure 1 of \citet{LeytonBrown:2014}. In particular,
satisfiable instances tend to be easier but show greater variation than unsatisfiable instances, and
there are exceptionally hard satisfiable instances \citep{Smith:1997}.

@@

What if we alter the edge probabilities for both the pattern graph and the target graph?  In the top
row of \cref{figure:non-induced} we show the phase transition for the non-induced variant, for
patterns of order 10, 20 and 30, a target of order 150, and varying pattern (x-axis) and target
(y-axis) edge probabilities. Each axis runs over 51 edge probabilities, from 0 to 1 in steps of
0.05; for each of these 2601 points, we generate ten random instances. The colour denotes (?? wrong
word?) the proportion of these instances which were found to be satisfiable.  Inside the orange
region, at the bottom right of each plot, every instance is unsatisfiable---here we are trying to
find a dense pattern in a sparse target. In the purple region, at the top left, every instance is
satisfiable---we are looking for a sparse pattern in a dense target (which is easy, since we only
have to preserve adjacency, not non-adjacency). The white band between the regions shows the
location of the phase transition: here, roughly half the instances are satisfiable. (We discuss the
black line below.)

On subsequent rows, we show the average number of search nodes used by the different algorithms. In
general, satisfiable instances are easy, until very close to the phase transition. As we hit the
phase transition and move into the unsatisfiable region, we see a complexity increase. Finally, as
we pass the phase transition and move deeper into the unsatisfiable region, instances become easier
again. This behaviour is largely solver-independent, although VF2 has a larger hard region than
Glasgow or LAD. Thus, although we have moved away from the single control parameter typically
studied, ?? I guess we want to say it's not that surprising?

@@

To predict the location of the phase transition, we can calculate the expected number of solutions
by making a few dubious assumptions involving rounding and independence. Since we are trying to find
an \emph{injective} mapping from a pattern $P = G(p, d_p)$ to a target $T = G(t, d_t)$, there are
$t^{\underline{p}} = t \cdot (t - 1) \cdot \ldots \cdot (t - p + 1)$ possible assignments of target
vertices to pattern vertices.  We expect the pattern to have $d_p \cdot \binom{p}{2}$ edges, so we
obtain the probability of all of these edges being mapped to edges in the target by raising $d_t$ to
this power, giving an expected number of solutions of \[ \langle Sol \rangle = t^{\underline{p}}
\cdot {d_t}^{d_p \cdot \binom{p}{2}} \textnormal{.} \] This formula predicts a very sharp phase
transition from $\langle Sol \rangle \ll 1$ to $\langle Sol \rangle \gg 1$. We plot where this
occurs using black lines in the first row of \cref{figure:non-induced}.

@@

?? Note that the size of the state space is defined differently to in binary CSPs: we're using the
all diff constraint to determine the size of the state space, and are not treating it as a
constraint. In other words, our generate-and-test approach is alldiff-aware. There are probably five
people in the world who care about this and would have seen it as very interesting twenty years ago.

@@

?? Point out how the value ordering heuristics, which were developed empirically, can be recovered
just by using this formula.  Note that the Glasgow heuristics are moving us towards the satisfiable
region, or increasing the expected number of solutions \citep{Gent:1996:EN}. LAD does SDF with no
tiebreaking, no value ordering heuristic.

@@

?? I suppose we should mention kappa \citep{Gent:1996:Kappa}, etc, here (or possibly in the
conclusion?). This whole thing is a bit subtle: maximising E(N) or Kappa as a principle is
encoding-independent, but whether this is done by SDF, deg, etc, is encoding-dependent. This seems
to be known by these same five people, but maybe isn't in the literature anywhere.

@@

In \cref{figure:induced} we repeat our experiments, finding induced isomorphisms. With a pattern of
order 10, we get two independent phase transitions (?? describe properly). For each of the solvers,
the satisfiable instances before either phase transition are generally easy, instances near the
phase transition are hard, and the large central unsatisfiable region is again easy.

?? For larger patterns of order 20 and 30, we have large unsatisfiable regions in the middle,
but these remain hard.

Pattern orders 14, 15 and 16 show the transition between the two styles.

@@

?? Note that the large unexpectedly hard region in the middle is solver-independent. Perhaps
subgraph isomorphism algorithms are not able to make use of adjacency and non-adjacency
simultaneously, and this is a weakness in how all three are designed? In the following section we
debunk this wonderful hypothesis.

?? Does kappa predict the hard region in the middle?

?? Maybe note a few solver-dependent hard instances.  There's odd unsat with LAD at the top right
(not EHPs), trivial for Glasgow: mapping a non-clique into a clique. Glasgow looks at non-edges and
can see immediately that no mapping is possible, but LAD doesn't do degree filtering on complements.

@@

?? Explain why we might guess degree heuristics just won't work for induced: for everything we say
about degree, the opposite holds too for the complement constraints.

?? Explain bottom row of \cref{figure:induced}, and give way of calculating it. The key point is
that depending upon where we are, changes to one of the two terms has more effect in the equation.

?? Possibly send the data to Lars, and get him to run his portfolios stuff on it, and then make fun
of machine learning?

@@

\subsection{Non-Random Instances}

@@

\begin{figure*}[t]
    \input{gen-graph-sat.tex}
    \caption{Behaviour of other solvers on the induced variant on smaller graphs, shown in the style of
        \cref{figure:non-induced}. The second row shows the number of search nodes used by the
    Glasgow algorithm, the third row shows the number of decisions made by the Clasp pseudo-boolean
solver, and the fourth row shows the number of search nodes used by BBMC on the clique
encoding.}\label{figure:alt}
\end{figure*}

?? We're looking at the odd stuff in the middle, which isn't near a phase transition, but is hard.

Is this just a weakness of current subgraph isomorphism solvers? In \cref{figure:alt} we repeat the
induced experiments on smaller pattern and target graphs, using different solving techniques.
Although these techniques are not competitive in absolute terms, we wish to see if the same pattern
of behaviour occurs.

Direct pseudo-boolean encoding: variables for each assignment, cardinality constraints, edge and
non-edge constraints. We use the Clasp solver. We see that hard instances remain hard, including
inside the central region, and that the easy satisfiable instances remain easy. ?? Nag James to
write this bit, and possibly kidnap his baby if he isn't cooperative.

@@

?? Clique encoding: association graph, a version of BBMC \citep{SanSegundo:2011} modified to solve the
decision problem. The hard instances in the central region remain hard. Additionally, some of the
easy satisfiable instances become hard. Possible weakness in variable ordering heuristics. The
interesting thing is here that clique knows nothing about adjacency or non-adjacency.

?? Some easy instances become hard, here, but hard instances remain hard. In particular, the large
central region for the larger pattern graphs which is not near the phase transitions is still
relatively difficult in all encodings. This suggests that these instances may genuinely be hard,
despite not being near a phase transition, and that this does not indicate a weakness in subgraph
isomorphism solvers.

@@

?? We found hard instances, some where we'd expect, some in unexpected places. These unexpected
instances, by and large, remained hard under reductions. We looked at how existing non-induced
heuristics could be recovered from Sol, and how this gives us better heuristics for the induced
variant.

?? Consider dynamic heuristics.

?? Other random models and non-random.

?? Does not give hard instances for graph isomorphism.

@@

\section*{Acknowledgements}

?? I suppose Craig could be given an inkling of undeserved gratitude if we have space left.
